---
title: "Life Expectancy Across Nations"
subtitle: "Project Proposal"
author: "Dressed to the 9's"
date: "October 30, 2019"
output: github_document
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(skimr)
library(gridExtra)
```

## Section 1. Introduction

Extensive research has been done to identify the many factors that influence life expectancy, such as demographic variables, income composition, and mortality rates. However, less research has surfaced identifying causes such as immunizations, which is a very important factor to consider when studying life expectancy due to the high incidence of infectious disease leading to mortality and morbidity. The World Health Organization (WHO) states that vaccination greatly reduces the burden of infectious diseases on a country itself. Therefore, reducing the burden of infectious diseases is important to increasing a country's average life expectancy.  

This dataset comes from the Global Health Observatory (GHO) data from the WHO. The data attempts to identify significant factors that may have an influence on a country's life expectancy, including information on nations’ immunization factors, mortality factors, economic factors, and other health-related factors. The dataset includes data from 193 countries across the years 2000- 2015; however, we have filtered the data to focus specifically on data from 183 countries within the year 2014 that may influence life expectancy. Life expectancy is reported in the data as the average age for each country. Therefore, countries can use this information to determine what is most important for them to focus on improving in order to improve overall life expectancy.  

Initially, we want to discover if there is a statistically significant difference between region and life expectancy. To investigate this potential relationship, we will create a new variable called 'region' to classify each country into its geographic region. We expect that we will see significant differences across regions, and if there are differences present, we will then investigate which regions have significantly different life expectancies than the others.  

Our aim is to build a model using multinomial regression that will identify which predictors from the WHO data are the most significant predictors of a nation’s life expectancy in the year 2014. We expect that predictors involving data on a nation’s immunizations (predictor variables Hepatitis B, HIV/AIDS, Polio, Diphtheria, and Measles) will be more significant predictors of life expectancy for the year 2014, rather than predictors involving mortality factors, economic factors, and other health-related factors, because these predictors relate directly to the burden of infectious diseases.  

However, there may be a difference between the significance of immunizations in developing countries versus developed countries where most of the population is already vaccinated. We expect that in developing countries, where there is a high burden of disease due to infectious diseases, immunizations will be more significant predictors of life expectancy than in developed countries, where the burden of disease is concentrated in non-communicable diseases rather than infectious diseases. To investigate this, we intend to explore the potential relationship between the life expectancies of countries classified as "developed" versus countries classified as "developing" using ANOVA, and we expect to find a significant difference between these two groups.   
  


## Section 2. Exploratory Data Analysis

```{r load-data, message=FALSE}
lifedata <- read_csv("data/life.csv")

lifedata <- lifedata%>%
  filter(year == 2014)
```

### Overall Description

The health-related data in this dataset was collected by the Global Health Observatory (GHO) data repository under World Health Organization (WHO). The economic data for each country was first reported on the United Nations website.

Before the merged dataset was reported on Kaggle, a few countries (such as Vanuatu, Tonga, Togo, and Cabo Verde) were removed because these observations were missing data for most of the variables.

While the Kaggle dataset included individual observations for each country for each year from 2000-2015, we will only explore the data for the year 2014 in order to narrow the focus and scope of our analysis.

Therefore, our final dataset will include 183 observations and 22 columns. Each observation represents a different world country for the year 2014.

### Response Variable

The response variable is the life expectancy of a country in years of age.

Its variable type is double.

```{r response_plot}
ggplot(data = lifedata, mapping = aes(x = life_expectancy))+
  geom_histogram(binwidth = 5)+
  labs(title = "Distribution of Life Expectancy", x = "life expectancy (in years)")
```

This variable is unimodal and centered approximtely around 75 years.

### Predictor Variables

There are 17 predictor variables that we are considering for our model. Status characterizes a country as developed or developing. Adult Mortality is the mortality rates of both sexes, the probability of dying between 15 and 60. Infant deaths is the number of infant deaths per 1000 population. Alcohol is the alcohol consumption, recorded per capita of 15+. Percentage expenditure is the expenditure on health as a percentage of GDP/capita, as a percentage. Hepatitis B is the HepB immunization coverage among 1 year olds, as a percentage. Measles is the number of reported cases per 1000 population. BMI is the average body mass index of the population. Under-five-deaths is the amount of under-five-deaths per 1000 population. Polio is the Pol3 immunization coverage among 1 year olds, as a percentage. Total expenditure is government expenditure on health as a percentage of total government expenditure. Diphtheria is the DTP3 immunization coverage among 1 year olds, as a percentage. HIV/AIDS is the deaths per 1000 live births with HIV/AIDS, from 0-4 years. GDP is the gross domestic product per capita, in dollars. Thinness 10-19 years is the prevalence of thinness among children age 10-19, as a percentage. Thinness 5-9 years is the prevalence of thinness among children age 5-9, as a percentage. Schooling is the number of years of schooling.

### Any Other Variables

There are a few variables that we are not using as predictors of life expectancy. Population of the country won't help us predict life expectancy. We are also not including income composition of recources because because this is the Human Development Index in terms of income composition of resources (index ranging from 0 to 1) and we do not expect this variable to have a significant effect on life expectancy.


### Preliminary EDA

````{r skim}
lifedata%>%
  skim()
```

As seen in the summary statistics, there are 183 observations and 22 variables. While some variables have some missing data, every country has a value for life expectancy, of which the average is 71.54 years.

We can first examine the status variable which indicates if a country is developed or developing:

```{r status}
ggplot(lifedata, aes(x = status))+
  geom_bar()+
  labs(title = "Count of Developed and Developing", x = "Status")

lifedata%>%
  count(status)
```

As seen in the bar graph above, there are many more developing countries than developed (151 vs. 32).

Additionally, we can examine boxplots of life expectancy for each of these two groups:

```{r boxplots}
ggplot(lifedata, aes(x = status, y = life_expectancy))+
  geom_boxplot()+
  labs(title = "Life Expectancy by Status", x = "Status",
       y = "Life Expectancy (in years)")
```

Preliminarily, these boxplots show that generally, developed countries have higher life expectancy than developing countries.

We would also like to examine immunization coverage for Hepatitis B, Polio, and Diphtheria. This is reported in terms of percent of 1-year-olds who are immunized.

```{r immunization_percent, warning=FALSE}
p1 <- ggplot(lifedata, aes(x = hepatitis_b))+
  geom_histogram(binwidth = 5)+
  labs(x = "Hepatitis B")

p2 <- ggplot(lifedata, aes(x = polio))+
  geom_histogram(binwidth = 5)+
  labs(x = "Polio")

p3 <- ggplot(lifedata, aes(x = diphtheria))+
  geom_histogram(binwidth = 5)+
  labs(x = "Diphtheria")

grid.arrange(p1, p2, p3, top = "Distributions of Immunization Coverage (%) ")
```

From these histograms it is evident that in most countries, most 1-year-olds are immunized for these diseases. However, for all the diseases there are still a few countries with almost a 0% immunization coverage, which likely affects life expectancy.

Another interesting variable records the number of under-five deaths per 1000 people in the population. This can be plotted against life expectancy to better understand the relationship between the two variables:

```{r under_five}

ggplot(lifedata, aes(x = under_five_deaths, y = life_expectancy))+
  geom_point()+
  labs(title = "Life Expectancy vs. Deaths Under 5", 
       x = "number of deaths under 5 years (per 1000 people)", 
       y = "life expectancy")

under_five_data <- lifedata%>%
  filter(under_five_deaths < 100)

ggplot(under_five_data, aes(x = under_five_deaths, y = life_expectancy))+
  geom_point()+
  labs(title = "Life Expectancy vs. Deaths Under 5", 
       subtitle = "for under_five_deaths < 100", 
       x = "number of deaths under 5 years (per 1000 people)", 
       y = "life expectancy")
```

Even when the number of deaths under age 5 is restricted to less than 100 (in the second plot), there surprisingly still seems to be a weak correlation between under_five_deaths and life_expectancy.

Perhaps an adult mortality rate will yield less surprising results when compared to life expectancy. This variable shows the mortality rate of both sexes (probability of dying between 15 and 60 years per 1000 population).

```{r adult_mortality}
ggplot(lifedata, aes(x = adult_mortality, y = life_expectancy))+
  geom_point()+
  labs(title = "Life Expectancy vs. Adult Mortality", 
       x = "number of deaths age 15-60 (per 1000 people)", 
       y = "life expectancy")+
  geom_smooth(method = lm)

mortality_data <- lifedata%>%
  filter(adult_mortality < 50)

ggplot(mortality_data, aes(x = adult_mortality, y = life_expectancy))+
  geom_point()+
  labs(title = "Life Expectancy vs. Adult Mortality", 
       subtitle = "for adult_mortality < 50", 
       x = "number of deaths age 15-60 (per 1000 people)", 
       y = "life expectancy")+
  geom_smooth(method = lm)
```

For adult mortality, as opposed to child mortality, two linear trends become obvious. One shows a negative correlation spanning from 0-500 adult deaths per 1000 people and the other is contained within the range of 0-50 adult deaths.

We can also examine government-related variables and their relationship with life expectancy, such as total_expenditure which measures the percent of total government expenditure spent on health:

```{r total_exp}
ggplot(lifedata, aes(x = total_expenditure, y = life_expectancy))+
  geom_point()+
  labs(title = "Life Expectancy vs. Percent of Total Expenditure",
       x = "% of total expenditure", y = "life expectancy (years)")
```

The relationship between these two variables appears to be weakly positive.



## Section 3. Regression Analysis Plan

### Interactions

In our regression analysis, we would like to first examine the relationship between region and life expectancy. We plan to create a new variable to classify each country into its geographic region, and then we will conduct an ANOVA test to discover if life expectancy varies across regions. If differences are present, we will utilize pairwise confidence intervals and confidence interval plots to determine which regions have significantly different life expectancies from others.  

Then, we expect that the status of a country (Developed vs. Developing) may also have an effect on life expectancy, as developing countries have less resources and are generally less stable than fully developed countries. We will perform a second ANOVA test to examine if there is a significant difference in life expectancy between developed and developing countries.  

Finally, we will investigate which country characteristics are the best predictors of life expectancy in a given country. First, we will create a new categorical variable to classify life expectancy: 48 years to 55 years will be considered poor, 55 years to 66 years will be considered fair, 66 years to 77 years will be considered good, and 77 years to 90 years will be considered excellent. We will use a multinomial logistic regression model to explore the effects of each country characteristic on the probability that the given country will have a certain category of life expectancy.  

We will begin by utilizing most variables in the dataset, including adult mortality, infant deaths, schooling, GDP, population, total expenditure, and BMI. We suspect that the influence of certain predictors, such as HIV/AIDS, polio, diptheria, and hepatitis B, may differ depending on geographic location (i.e. life expectancy in countries with readily available vaccine supplies may not be as affected by these diseases as life expectancy in countries without readily available vaccines). To address this, when fitting our model, we will calculate the interaction effect of region on HIV/AIDS, polio, diptheria, and hepatitis B.  

### Model Selection

To create and improve our model, we will use backwards selection based on Schwarz's Bayesian Information Criterion, or BIC. We will begin with all possible variables that we are considering as potentially significant to life expectancy category, status, adult mortality, infant deaths, alcohol, percentage expenditure, hepatitis B, measles, BMI, under-five-deaths, polio, total expenditure, diphtheria, HIV/AIDS, GDP, thinness 10-19 years, thinness 5-9 years, and schooling. We will continually remove variables and select the model with the lowest BIC. We selected this criterion because the BIC possesses a larger penalty for unnecessary predictors than the AIC and adjusted R^2^, so it will result in a more parsimonious model. To make our model more useful and easier to understand, we want it to possess a lower number of significant predictors, which is accomplished by the BIC.

### Modeling Technique  

To create our model, we will use multinomial regression to display the effects of country characteristics on the odds that a country will have a certain category of life expectancy (poor, fair, good, or excellent). We chose this technique because we wanted to see the effects of multiple variables on a categorical response with several levels -- this could not be accomplished by simple linear regression, multiple linear regression, or logistic regression. We believe that when investigating countries that are in need of global assistance, the exact life expectancy is not required -- rather, simply knowing if the life expectancy is very low/high is sufficient for recognizing a struggling nation. We believe multinomial linear regression will allow us to utilize quantitative characteristics of different countries to predict an important categorical response.  

We will use ANOVA for preliminary investigation of whether life expectancy differs between regions. ANOVA allows us to see if a response variable is generally equal across categories or if it differs, and if life expectancy varies across regions, this is a good indicator to study other characteristics of region with high and low life expectancies to see what may be influencing life expectancy. We are also using the interaction effect technique to discover how differences in region may affect the influence of certain diseases on a country's life expectancy, as this technique allows us to control the influence (coefficients) of certain diseases in our model if it is significantly different over regions.


## Section 4. References

Help with the gridExtra package and grid.arrange() function:

https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html#title-andor-annotations

Introduction: 

WHO | Vaccination greatly reduces disease, disability, death and inequity worldwide. (n.d.). Retrieved October 29, 2019, from WHO website: https://www.who.int/bulletin/volumes/86/2/07-040089/en/

## The Data

https://www.kaggle.com/kumarajarshi/life-expectancy-who 

```{r glimpse_data}
glimpse(lifedata)
```

